Bayesian Sparsification of Deep $\cplx$-valued networks

### 2020-03-21 paper/paper.tex
thorough proofread 4 (scattered) * merged sub 2.1 into sec 2 * improved names of c-valued dropout section, and rewritten its intro * reworded paragraph on non-overlapping patches * moved the first piece from `priors` sub to `c-gaussian` sub * scattered edits

The paper is structured as follows. In sec.~\ref{sec:bayesian_dropout} we review Bayesian
dropout techniques, and in sec.~\ref{sec:c_valued_networks} we provide a brief summary of
the inner working of complex-valued networks as functions of complex argument. The main
contribution of this study is presented in sec.~\ref{sec:dropout_for_c_valued_layers},
where we consider different variational approximations and priors, outline the tricks
and derive the penalty terms in the evidence lower bound. In sec.~\ref{sec:experiments}
we evaluate the sparsification rate, compare the resulting performance of various $\cplx$-networks
proposed in prior work, and discuss the outcomes.

@@

\section{Bayesian Dropout} % (fold)
\label{sec:bayesian_dropout}

@@

\subsection{Variational Dropout} % (fold)
\label{sub:variational_dropout}

@@

% subsection dropout (end)

@@

\section{Dropout for $\cplx$-valued layers} % (fold)
\label{sec:dropout_for_c_valued_layers}

In this section we review the complex Gaussian distribution and discuss its properties. Then
we introduce factorized dropout posterior for complex-valued parameters and the relevant
version of the local reparameterization trick. Then we move on to choosing suitable priors
and then to deriving explicit expressions for the divergence penalties in \eqref{eq:elbo}
in terms of \verify{special} functions.

\subsection{Complex Gaussians and Local Reparameterization} % (fold)
\label{sub:c_gauss_and_local_rep}

@@

The $\cplx$-variant of the local reparameterization trick utilizes the fact that Gaussian
vectors are closed under affine transformations:
% $\cplx$-Gaussians are closed under affine transformations
\begin{equation}  \label{eq:cn-affine}
  b + A z \sim \mathcal{N}_n^{\cplx}\bigl(
      b + A\mu, A \Gamma A^{\hop}, A C A^\top
    \bigr)
  \,,
\end{equation}
for $A \in \cplx^{n \times m}$ and $b \in \cplx^{n}$. Therefore, if $W \in \cplx^{n\times m}$
is drawn from a fully factorized approximation
\begin{equation}  \label{eq:c-gauss-vi-general}
  W \sim q(W)
    = \prod_{ij} \mathcal{N}^{\cplx} \bigl(
      w_{ij} \vert\, \mu_{ij}, \Sigma_{ij}, C_{ij}
    \bigr)
  \,,
\end{equation}
with $
  \mu, C \in \cplx^{n\times m}
$, $\Sigma \in \real^{n\times m}$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$, then for $x \in \cplx^m$ and $b \in \cplx^n$ the vector $y = b + W x$ has independent
components with
\begin{equation}  \label{eq:cplx-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \Bigl(
        % e_i^\top \mu x + b_i,
        b_i + \sum_{j=1}^m \mu_{ij} x_j,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \Bigr)
    \,.
\end{equation}
The trick requires three operations, i.e. $b + \mu x$, $\Sigma \lvert x \rvert^2$ and
$C x^2$ with the complex modulus and square applied elementwise, and can be applied to
any layer, the output of which depends linearly on its parameters, such as convolutional,
affine, and bilinear transformations ($
  (x, z) \mapsto x^\top W^{(j)} z + b_j
$). Similar to $\real$ case, Variational Dropout for $\cplx$ convolutions draws independent
realizations of $W$ for each spatial patch in the input \citep{molchanov_variational_2017}.
This provides faster computations and better statistical efficiency of the SGVB gradient
estimator stemming from decorrelated gradient estimates from overlapping patches
\citep{kingma_variational_2015}, allowing \eqref{eq:cplx-gauss-trick} to leverage $\cplx$
convolutions of the relation and variance kernels with elementwise complex squares $x^2$
and amplitudes $\lvert x \rvert^2$.
% Weight sharing is essentially pushed to shared variational parameters $\theta$ in the approximation $q_\theta$.
% A convolution is a matrix product of $W$, unrolled into a toeplitz matrix, and im-to-col
% flattened intput $x$

@@

\subsection{The choice of priors and Kullback-Leibler divergence} % (fold)
\label{sub:priors_and_kullback_leibler_divergence}

For $\cplx$-Bayesian Dropout we propose to use fully factorized proper $\cplx$-Gaussian
approximation \eqref{eq:c-gauss-vi-general} for weights and point estimates of biases in
dense linear, convolutional and other, effectively, affine layers, i.e. we let $
  \Sigma_{ij} = \alpha_{ij} \lvert \mu_{ij} \rvert^2
$ and fix $C_{ij} = 0$. Under this approximation the relation coefficients in \eqref{eq:cplx-gauss-trick}
are zero.
%
For a fully factorized approximation $q(\omega)$ and factorized prior belief $
  \pi(\omega) = \prod_{ij} \pi(\omega_{ij})
$,
the divergence term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q \| \pi)
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = - \sum_{ij}
        \mathbb{H}(q(\omega_{ij}))
        + \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
    \,.
\end{equation}
We consider two fully factorized priors: an improper prior for Variational Dropout and
$\cplx$-Gaussian ARD prior. We omit subscripts ${ij}$ for brevity in the next sections.

@@

% section dropout_for_c_valued_layers (end)

@@

% write a note on the incompressible parameters (bias, batch norm etc.)
The compression rate is computed based on the number of IEEE754 floating point values
needed to store the model and equals $
  \tfrac{n_\mathtt{par}}{n_\mathtt{par} - n_\mathtt{zer}}
$, where $n_\mathtt{zer}$ is the number of explicit zeros, $n_\mathtt{par}$ is the total
number of floats. In particular, each parameter is counted once in a $\real$-valued network,
but twice in a $\cplx$-network. Each model has a compression limit, determined by its
layers' biases, shift and scaling from $\real$-and $\cplx$-batch normalization.

@@

% on compresssion
Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sub:variational_dropout}).
Since the $q_\theta$ factorizes into univariate distributions, a $\cplx$ or $\real$ parameter
is considered non-zero if $\log \alpha \leq \tau$ for $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$, or zero otherwise.

>>>>

The paper is structured as follows. In sec.~\ref{sec:variational_dropout} we review Bayesian
dropout techniques, and in sec.~\ref{sec:c_valued_networks} we provide a brief summary of
the inner working of complex-valued networks as functions of complex argument. The main
contribution of this study is presented in sec.~\ref{sec:c_variational_dropout},
where we consider different variational approximations and priors, outline the tricks
and derive the penalty terms in the evidence lower bound. In sec.~\ref{sec:experiments}
we evaluate the sparsification rate, compare the resulting performance of various $\cplx$-networks
proposed in prior work, and discuss the outcomes.

@@

\section{Variational Dropout} % (fold)
\label{sec:variational_dropout}

@@

@@

@@

\section{$\cplx$-Variational Dropout} % (fold)
\label{sec:c_variational_dropout}

We propose to use fully factorized complex Gaussian posterior approximation and $\cplx$
variant of the local reparameterization trick to extend Bayesian Dropout techniques to
$\cplx$-valued networks. In this section we describe the approximation and derive the
divergence penalties in \eqref{eq:elbo}.

\subsection{$\cplx$-Gaussian Distribution} % (fold)
\label{sub:c_gauss_and_local_rep}

@@

The $\cplx$-variant of the local reparameterization trick utilizes the fact that Gaussian
vectors are closed under affine transformations:
% $\cplx$-Gaussians are closed under affine transformations
\begin{equation}  \label{eq:cn-affine}
  b + A z \sim \mathcal{N}_n^{\cplx}\bigl(
      b + A\mu, A \Gamma A^{\hop}, A C A^\top
    \bigr)
  \,,
\end{equation}
for $A \in \cplx^{n \times m}$ and $b \in \cplx^{n}$. Therefore, if $W \in \cplx^{n\times m}$
is drawn from a fully factorized approximation
\begin{equation}  \label{eq:c-gauss-vi-general}
  W \sim q(W)
    = \prod_{ij} \mathcal{N}^{\cplx} \bigl(
      w_{ij} \vert\, \mu_{ij}, \Sigma_{ij}, C_{ij}
    \bigr)
  \,,
\end{equation}
with $
  \mu, C \in \cplx^{n\times m}
$, $\Sigma \in \real^{n\times m}$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$, then for $x \in \cplx^m$ and $b \in \cplx^n$ the vector $y = b + W x$ has independent
components with
\begin{equation}  \label{eq:cplx-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \Bigl(
        % e_i^\top \mu x + b_i,
        b_i + \sum_{j=1}^m \mu_{ij} x_j,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \Bigr)
    \,.
\end{equation}
The trick requires three operations, i.e. $b + \mu x$, $\Sigma \lvert x \rvert^2$ and
$C x^2$ with the complex modulus and square applied elementwise, and can be applied to
any layer, the output of which depends linearly on its parameters, such as convolutional,
affine, and bilinear transformations ($
  (x, z) \mapsto x^\top W^{(j)} z + b_j
$). Similar to $\real$ case, Variational Dropout for $\cplx$ convolutions draws independent
realizations of $W$ for each spatial patch in the input \citep{molchanov_variational_2017}.
This provides faster computations and better statistical efficiency of the SGVB gradient
estimator by eliminating correlation from overlapping patches \citep{kingma_variational_2015}
and allowing \eqref{eq:cplx-gauss-trick} to leverage $\cplx$ convolutions of the relation
and variance kernels with elementwise complex squares $x^2$ and amplitudes $\lvert x \rvert^2$.
% Weight sharing is essentially pushed to shared variational parameters $\theta$ in the approximation $q_\theta$.
% A convolution is a matrix product of $W$, unrolled into a toeplitz matrix, and im-to-col
% flattened intput $x$

For $\cplx$-Bayesian Dropout we propose to use fully factorized proper $\cplx$-Gaussian
approximation \eqref{eq:c-gauss-vi-general} for weights and point estimates of biases in
dense linear, convolutional and other, effectively, affine layers, i.e. $
  \Sigma_{ij} = \alpha_{ij} \lvert \mu_{ij} \rvert^2
$ and $C_{ij} = 0$. This forces the relation coefficients in \eqref{eq:cplx-gauss-trick}
to be zero.

@@

\subsection{The priors} % (fold)
\label{sub:the_priors}

For a fully factorized approximation $q(\omega)$ and factorized prior belief $
  \pi(\omega) = \prod_{ij} \pi(\omega_{ij})
$, the divergence term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q \| \pi)
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = - \sum_{ij}
        \mathbb{H}(q(\omega_{ij}))
        + \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
    \,.
\end{equation}
We consider two fully factorized priors: an improper prior for Variational Dropout and
$\cplx$-Gaussian ARD prior. We omit subscripts ${ij}$ for brevity in the next sections.

@@

% section c_variational_dropout (end)

@@

% write a note on the incompressible parameters (bias, batch norm etc.)
The compression rate is computed based on the number of IEEE754 floating point values
needed to store the model and equals $
  \tfrac{n_\mathtt{par}}{n_\mathtt{par} - n_\mathtt{zer}}
$, where $n_\mathtt{zer}$ is the number of explicit zeros, $n_\mathtt{par}$ is the total
number of floats. In particular, each parameter is counted once in a $\real$-valued network,
but twice in a $\cplx$-network. Each model has a compression limit, determined by its
layers' biases, shift and scaling from $\real$- and $\cplx$-valued batch normalization.

@@

% on compresssion
Between \textit{``sparsify''} and \textit{``fine-tune''} stages we compute masks of non-zero
weights in each layer based on the relevance scores $\alpha$ (sec.~\ref{sec:variational_dropout}).
Since the $q_\theta$ factorizes into univariate distributions, a $\cplx$ or $\real$ parameter
is considered non-zero if $\log \alpha \leq \tau$ for $
  \alpha = \tfrac{\sigma^2}{\lvert\mu\rvert^2}
$, or zero otherwise.
